\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-02-07 fluB.tex
Lineage should now only refer to Vic/Yam.
Addressing #7.
Also adding alternative Vic/Yam colour scheme (orange/gray) to the
current one (orange/green). If everyone is happy with the new one I'll
update all 3 figures with the new one.

\begin{figure}[h]
	\centering		
	\includegraphics[width=0.75\textwidth]{figures/InfB_tmrcaOT_lines.png}
	\caption{\textbf{Oldest TMRCA of all surviving lineages over time.}
PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turn over.
PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
Each point is the TMRCA of all surviving lineages existing at each time slice through the tree and vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
	\label{tmrcaOT}
\end{figure}

@@

\begin{figure}[h]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/RailroadPlotDated.pdf}
	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored red and lineages that derive from the original Yamagata clade colored blue.
Inter-lineage reassortment events are indicated by lines entering a different genome.
The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
Lineage extinction dates are not shown accurately.
\tbc{Giving this a bit more vertical spread (increasing aspect ratio) would likely clarify things a bit.}}
	\label{railroadPlot}
\end{figure}

@@

Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome, i.e.\ that have shared a recent common virion.
In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \cite{turner2005}.

@@

Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (see Supplementary information).
The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
However, our analyses do not indicate any obvious or dramatic differences between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two.

@@

\subsection*{Conclusion}
%The $\Delta$TMRCA method, which is an extension of patristic (or phylogenetic) distance methods, can be applied to cases where reticulate evolution is suspected and phylogenetic trees can readily be made.
%Patristic distance methods are far from new, although their use in conjunction with temporally calibrated phylogenies has been limited.
For example, one study focused on patristic (or phylogenetic) distances to predict viral titer in \textit{Drosophila} species infected by viruses of other \textit{Drosophila} species \cite{longdon2011}.
Another study has used the equivalent statistic to our $\Delta$TMRCA (called $\Delta$\textit{t}) to assess the temporal incongruence in phylogenies of 171 amphibian species induced by using different calibrations \cite{ruane2011}.
%By using relatively straightforward measures of diversity, such as TMRCA of all surviving lineages over time and mean pairwise TMRCA between labels...

>>>>

\begin{figure}[h]
	\centering		
	\includegraphics[width=0.75\textwidth]{figures/InfB_tmrcaOT_lines.png}
	\caption{\textbf{Oldest TMRCA of all surviving branches over time.}
PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turnover.
PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
Each point is the TMRCA of all surviving lineages existing at each time slice through the tree and vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
	\label{tmrcaOT}
\end{figure}

@@

\begin{figure}[h]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/RailroadPlotDated.pdf}
	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored red and lineages that derive from the original Yamagata clade colored blue.
Inter-lineage reassortment events are indicated by lines entering a different genome.
The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
Lineage extinction dates are not shown accurately.
	\label{railroadPlot}
\end{figure}

@@

Reassortment `distances' represent the temporal separation of reassorting branches and suggests that reassortments between PB1, PB2 and HA segments tend to combine branches that have relatively recently been part of the same genome, i.e.\ that have shared a recent common virion.
In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \cite{turner2005}.

@@

Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (see Supplementary information).
The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
However, our analyses do not indicate any obvious or dramatic differences between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (see Supplementary information).

@@

\subsection*{Conclusion}
%The $\Delta$TMRCA method, which is an extension of patristic (or phylogenetic) distance methods, can be applied to cases where reticulate evolution is suspected and phylogenetic trees can readily be made.
%Patristic distance methods are far from new, although their use in conjunction with temporally calibrated phylogenies has been limited.
For example, one study focused on patristic (or phylogenetic) distances to predict viral titer in \textit{Drosophila} species infected by viruses of other \textit{Drosophila} species \cite{longdon2011}.
Another study has used the equivalent statistic to our $\Delta$TMRCA (called $\Delta$\textit{t}) to assess the temporal incongruence in phylogenies of 171 amphibian species induced by using different calibrations \cite{ruane2011}.
%By using relatively straightforward measures of diversity, such as TMRCA of all surviving branches over time and mean pairwise TMRCA between labels...
